---
title: "Speed comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

# Purpose

FastSurvival is designed for repeated evaluation inside large simulation
loops. This vignette shows how to benchmark each function against its
standard counterpart and reports representative results. The benchmark code
is shown but not executed when the vignette is built, because timing many
microbenchmark replicates would exceed the build-time limits. To reproduce
the numbers, run the code blocks interactively.

The reported figures are median times from microbenchmark replicates on a
single desktop machine. Absolute timings depend on hardware, sample size, and
event rate, so the ratios matter more than the raw values.

```{r load}
library(FastSurvival)
library(survival)
library(microbenchmark)
# Comparison packages used in the benchmarks below
library(survRM2)
library(nph)
```

# Setup

The key to the speed gain is that the analysis functions accept pre-sorted
vectors. Inside a simulation loop the data are sorted once and reused, so the
sort cost is paid a single time rather than on every call. We build a single
phase-3-sized dataset for the benchmarks.

```{r data}
set.seed(1)
n  <- 600
tt <- rexp(n, rate = 0.05)
ev <- rbinom(n, 1, 0.8)
gp <- rep(1:2, each = n / 2)

ord <- order(tt)
t_s <- tt[ord]
e_s <- ev[ord]
g_s <- gp[ord]
```

# survfit_fast vs survfit + summary

```{r bench-survfit}
microbenchmark(
  fast = survfit_fast(t_s, e_s, t_eval = 20, presorted = TRUE),
  base = summary(survfit(Surv(tt, ev) ~ 1), times = 20),
  times = 1000
)
```

# survdiff_fast vs survdiff

```{r bench-survdiff}
microbenchmark(
  fast = survdiff_fast(t_s, e_s, g_s, control = 1, side = 2, presorted = TRUE),
  base = survdiff(Surv(tt, ev) ~ gp),
  times = 1000
)
```

# coxph_fast vs coxph

```{r bench-coxph}
microbenchmark(
  fast = coxph_fast(t_s, e_s, g_s, control = 1, presorted = TRUE),
  base = coxph(Surv(tt, ev) ~ I(gp == 2)),
  times = 1000
)
```

# rmst_fast vs survRM2::rmst2

```{r bench-rmst}
arm <- as.integer(gp == 2)
microbenchmark(
  fast = rmst_fast(t_s, e_s, g_s, control = 1, tau = 20, presorted = TRUE),
  base = survRM2::rmst2(time = tt, status = ev, arm = arm, tau = 20),
  times = 1000
)
```

# survdiff_fast(weight = "fh") vs nph::logrank.test

```{r bench-wlr}
microbenchmark(
  fast = survdiff_fast(t_s, e_s, g_s, control = 1, side = 2,
                       weight = "fh", rho = 0, gamma = 1, presorted = TRUE),
  base = nph::logrank.test(tt, ev, gp, rho = 0, gamma = 1),
  times = 1000
)
```

# Representative results

The table below summarizes representative median timings on a typical
phase-3 dataset (n = 600, event rate 80%) with `presorted = TRUE`. The exact
values will differ on your machine, but the order of magnitude of the speedup
is stable.

| Function | Replaces | Approximate speed gain |
|----------|----------|------------------------|
| `survfit_fast()` | `survfit()` + `summary()` at one time point | ~70x |
| `survdiff_fast()` | `survdiff()` | ~50x |
| `coxph_fast()` | `coxph()` (point estimate + Wald CI) | ~50x |
| `rmst_fast()` | `survRM2::rmst2()` | ~60x |
| `survdiff_fast(weight = "fh")` | `nph::logrank.test()` | ~500x |

# Why it is faster

Each function avoids the overhead that the standard implementations incur on
every call. The standard functions parse a formula, build an S3 model object,
and construct intermediate vectors before producing the result, which is
appropriate for interactive use but wasteful when the same operation is
repeated thousands of times. The FastSurvival functions take plain vectors,
do the core computation in a single C++ pass over the data, and return a
lightweight numeric vector. When the input is already sorted the sort cost is
avoided entirely. In a simulation loop these savings accumulate across every
iteration.

# References

Homma, G. (2025). One step from Pike to Cox: a closed-form hazard ratio
estimator. *Manuscript under review.*

Collett, D. (2014). *Modelling Survival Data in Medical Research* (3rd ed.).
Chapman and Hall/CRC.
