---
title: "Validation of FastSurvival"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation of FastSurvival}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
have_survival <- requireNamespace("survival", quietly = TRUE)
have_survRM2  <- requireNamespace("survRM2", quietly = TRUE)
have_survAH   <- requireNamespace("survAH", quietly = TRUE)
have_nph      <- requireNamespace("nph", quietly = TRUE)
have_nphRCT   <- requireNamespace("nphRCT", quietly = TRUE)
```

# Purpose

FastSurvival is built for speed, but a fast estimator is only useful if it
returns the same answer as the established implementation. This vignette
checks the numerical agreement between each FastSurvival function and a
reference from a well-known package: the survival package for the
Kaplan-Meier estimate, the log-rank test, the Cox hazard ratio, and milestone
survival; survRM2 for the restricted mean survival time; and survAH for the
average hazard with survival weight. The weighted log-rank and max-combo tests
are checked against the nph package, and the robust modestly-weighted test
against the nphRCT package. The same comparisons form the basis of
the automated test suite shipped with the package.

```{r load}
library(FastSurvival)
```

# Kaplan-Meier survival

`survfit_fast()` evaluates the Kaplan-Meier estimate at a single time point.
We compare against `summary(survfit(...))` at the same time on the `ovarian`
dataset.

```{r km, eval = have_survival}
library(survival)

ord <- order(ovarian$futime)
t_s <- ovarian$futime[ord]
e_s <- ovarian$fustat[ord]

fast <- survfit_fast(t_s, e_s, t_eval = 500, conf.type = "log")

fit  <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
ref  <- summary(fit, times = 500)

data.frame(
  quantity = c("survival", "std.err"),
  fast     = c(unclass(fast)["surv"], unclass(fast)["std.err"]),
  survival = c(ref$surv, ref$std.err),
  row.names = NULL
)
```

# Log-rank test

`survdiff_fast()` returns a two-sided chi-square statistic when `side = 2`,
matching the statistic from `survdiff()`.

```{r logrank, eval = have_survival}
fast_lr <- survdiff_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                         control = 1, side = 2)

ref_lr  <- survdiff(Surv(futime, fustat) ~ rx, data = ovarian)

c(fast = as.numeric(fast_lr), survival = ref_lr$chisq)
```

# Weighted log-rank test

`survdiff_fast()` also computes Fleming-Harrington G(rho, gamma) weighted
log-rank tests through the `weight = "fh"` argument. The
[nph](https://cran.r-project.org/package=nph) package provides
`logrank.test()` with the same family, so we compare the chi-square statistic
across several weight choices on the `ovarian` data.

```{r wlr, eval = have_survival && have_nph}
fh_grid <- data.frame(rho = c(0, 1, 0, 1), gamma = c(1, 0, 0, 1))

do.call(rbind, lapply(seq_len(nrow(fh_grid)), function(i) {
  r <- fh_grid$rho[i]
  g <- fh_grid$gamma[i]
  fast <- as.numeric(survdiff_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                                   control = 1, side = 2,
                                   weight = "fh", rho = r, gamma = g))
  nph_chisq <- nph::logrank.test(ovarian$futime, ovarian$fustat, ovarian$rx,
                                 rho = r, gamma = g)$test$Chisq
  data.frame(rho = r, gamma = g, fast = fast, nph = nph_chisq)
}))
```

The two implementations agree to numerical precision across the weight family,
including the ordinary log-rank test recovered at `rho = 0, gamma = 0`.

# Cox hazard ratio

`coxph_fast()` returns the Pike-Halley Estimator, a closed-form approximation
to the Cox partial likelihood maximizer. On data with a moderate hazard ratio
the two agree closely. We report the log hazard ratio from both.

```{r cox, eval = have_survival}
fast_hr <- coxph_fast(ovarian$futime, ovarian$fustat, ovarian$rx, control = 1)

ref_cox <- coxph(Surv(futime, fustat) ~ I(rx == 2), data = ovarian)

c(fast = unclass(fast_hr)["coef"], cox = unname(coef(ref_cox)))
```

# Restricted mean survival time

`rmst_fast()` integrates the Kaplan-Meier survival curve up to a horizon. We
compare the two-group RMST difference against `survRM2::rmst2()`.

```{r rmst, eval = have_survival && have_survRM2}
library(survRM2)

arm  <- as.integer(ovarian$rx == 2)
tau  <- 500

fast_rmst <- rmst_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                       control = 1, tau = tau)

ref_rmst  <- rmst2(time = ovarian$futime, status = ovarian$fustat,
                   arm = arm, tau = tau)

c(fast = unclass(fast_rmst)["diff"],
  survRM2 = ref_rmst$unadjusted.result[1, 1])
```

# Milestone survival

`milestone_fast()` compares Kaplan-Meier survival between two groups at a
milestone timepoint. The per-group survival probabilities match those from
`survfit()`.

```{r milestone, eval = have_survival}
tstar <- 500

fast_ms <- milestone_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                          control = 1, tau = tstar, method = "wald")

fit_g <- survfit(Surv(futime, fustat) ~ rx, data = ovarian)
ref_g <- summary(fit_g, times = tstar)

ms_surv <- fast_ms$surv

data.frame(
  group    = c("control", "treatment"),
  fast     = c(ms_surv["control"], ms_surv["treatment"]),
  survival = ref_g$surv,
  row.names = NULL
)
```

# Average hazard with survival weight

`ahsw_fast()` computes the average hazard with survival weight of Uno and
Horiguchi. We compare the per-group average hazard against `survAH::ah2()`.

```{r ahsw, eval = have_survival && have_survAH}
library(survAH)

arm <- as.integer(ovarian$rx == 2)
tau <- 500

fast_ah <- ahsw_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                     control = 1, tau = tau)

ref_ah  <- ah2(time = ovarian$futime, status = ovarian$fustat,
               arm = arm, tau = tau)

data.frame(
  quantity = c("AH (control)", "AH (treatment)"),
  fast     = unclass(fast_ah)[c("ah.ctrl", "ah.trt")],
  survAH   = c(ref_ah$ah["AH (arm0)", "Est."],
               ref_ah$ah["AH (arm1)", "Est."]),
  row.names = NULL
)
```

The average hazard is the ratio of the cumulative event probability to the
restricted mean survival time, so for the `ovarian` data, where follow-up is
measured in days, the values are on the order of 1e-03 per day in both groups.

# Average hazard ratio

`ahr_fast()` computes the Kalbfleisch-Prentice average hazard ratio between
two groups over a restricted interval. The reference implementation is
`ahrKM()` from the AHR package, but that package has been archived on CRAN, so
here we check the point estimates against a direct survival-based computation:
each group's Kaplan-Meier curve from `survfit()`, integrated to form the group
shares of the total hazard.

```{r ahr, eval = have_survival}
tau <- 500

fast_ahr <- ahr_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                     control = 1, tau = tau)

g  <- ovarian$rx
ev <- c(ovarian$futime[g == 1 & ovarian$fustat == 1],
        ovarian$futime[g == 2 & ovarian$fustat == 1])
grid <- sort(unique(c(0, ev[ev <= tau], tau)))

km_on_grid <- function(gi) {
  sf <- survfit(Surv(futime, fustat) ~ 1, data = ovarian[g == gi, ])
  approxfun(sf$time, sf$surv, method = "constant",
            yleft = 1, rule = 2, f = 0)(grid)
}

S1  <- km_on_grid(1)
S2  <- km_on_grid(2)
m   <- length(grid)
dS1 <- S1 - c(1, S1[-m])
GL  <- S1[m] * S2[m]
ref_theta1 <- -sum(S2 * dS1) / (1 - GL)

data.frame(
  quantity  = c("theta (group 1)", "theta (group 2)", "AHR"),
  fast      = c(fast_ahr$theta[[1]], fast_ahr$theta[[2]], fast_ahr$ahr),
  reference = c(ref_theta1, 1 - ref_theta1, (1 - ref_theta1) / ref_theta1),
  row.names = NULL
)
```

The point estimates match the survival-based reference to numerical precision.
The average hazard ratio was additionally cross-checked against the archived
AHR package, the reference implementation used by Dormuth et al. (2024): the
two group shares, the average hazard ratio, the variances, and both test
statistics agree to within the order of 1e-15. That external comparison is
reproducible with the `tools/verify_ahr_fast.R` script shipped in the package
sources.

# Max-combo test

`maxcombo_fast()` computes the max-combo test, the most extreme of a set of
Fleming-Harrington weighted log-rank statistics. The `nph` package provides
`logrank.maxtest()`, whose default weight set is FH(0, 0), FH(0, 1), and
FH(1, 0). We request the same three weights from `maxcombo_fast()` and compare
the component Z-scores. The individual Z-scores are kept in the `z` attribute,
signed so that a negative value indicates benefit for the treatment group
under the package convention. Because `nph` orients the contrast in the
opposite direction, the signs are mirrored, so we compare absolute values.

```{r maxcombo, eval = have_survival && have_nph}
rho   <- c(0, 0, 1)
gamma <- c(0, 1, 0)

fast_mc <- maxcombo_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                         control = 1, side = 2, rho = rho, gamma = gamma)

nph_mc <- nph::logrank.maxtest(ovarian$futime, ovarian$fustat, ovarian$rx)

data.frame(
  weight = c("FH(0,0)", "FH(0,1)", "FH(1,0)"),
  fast   = abs(attr(fast_mc, "z")),
  nph    = abs(nph_mc$tests$z),
  row.names = NULL
)
```

The component statistics agree in absolute value. The final max-combo
statistic is the largest of these, which for the two-sided test is
`max(abs(Z))` and is therefore always non-negative. The p-values from the two
packages are close but not identical, since they use different numerical
methods for the joint distribution (a multivariate normal integral here, a
Bonferroni-style combination in `nph`).

# Robust modestly-weighted log-rank test

`rmw_fast()` computes the robust modestly-weighted (rMW) test of Magirr and
Ohrn, the maximum of the standard log-rank statistic and a single
modestly-weighted log-rank statistic. Each component is a weighted log-rank
test, so we check the two component Z-scores against
[nphRCT](https://cran.r-project.org/package=nphRCT), the implementation used
by the method's authors. The modestly-weighted component uses the
survival-threshold parameterization, so `s_star = 1` recovers the standard
log-rank test and `s_star = 0.5` gives the modestly-weighted component. As
with the max-combo test the package orients a negative Z toward treatment
benefit while `nphRCT` uses the opposite sign, so we compare absolute values.

```{r rmw, eval = have_survival && have_nphRCT}
ov_df <- data.frame(
  time  = ovarian$futime,
  event = ovarian$fustat,
  arm   = factor(ifelse(ovarian$rx == 1, "control", "experimental"),
                 levels = c("control", "experimental"))
)

fit_rmw <- rmw_fast(ov_df$time, ov_df$event, ov_df$arm,
                    control = "control", side = 1, s_star = 0.5)

z_lr_nph <- nphRCT::wlrt(Surv(time, event) ~ arm, data = ov_df,
                         method = "mw", s_star = 1)$z
z_mw_nph <- nphRCT::wlrt(Surv(time, event) ~ arm, data = ov_df,
                         method = "mw", s_star = 0.5)$z

data.frame(
  component = c("log-rank (s_star = 1)", "modestly-weighted (s_star = 0.5)"),
  fast      = abs(attr(fit_rmw, "z")),
  nphRCT    = abs(c(z_lr_nph, z_mw_nph)),
  row.names = NULL
)
```

The component statistics agree in absolute value. The null correlation of the
two components, reported in the `corr` attribute, matches the covariance
computed by the authors' `find_cor` routine to numerical precision (0.98 on
the `ovarian` data). The combined statistic and one-sided p-value then follow
from the bivariate normal distribution of the two components, evaluated with
`mvtnorm::pmvnorm`. On the POPLAR overall-survival data analyzed in the
original article, `rmw_fast()` reproduces the published one-sided p-values of
0.0028 for the log-rank test, 0.0009 for the modestly-weighted test, and
0.0012 for the rMW test, with a component correlation of 0.97.

# Summary

Across all functions the FastSurvival results reproduce the reference values
from the established packages. The point estimates and test statistics agree
to numerical precision for the Kaplan-Meier, log-rank, RMST, milestone, and
average-hazard quantities, the average hazard ratio matches a survival-based
reference, and the closed-form Cox hazard ratio agrees with
the partial-likelihood maximizer to the order expected for the Pike-Halley
approximation. This agreement is verified continuously by the package test
suite.

# References

Tang, Y. (2021). Some new confidence intervals for Kaplan-Meier based
estimators from one and two sample survival data. *Statistics in Medicine*,
40(23), 4961-4976.

Uno, H., & Horiguchi, M. (2023). Ratio and difference of average hazard with
survival weight: new measures to quantify survival benefit of new therapy.
*Statistics in Medicine*, 42(7), 936-952.

Kalbfleisch, J. D., & Prentice, R. L. (1981). Estimation of the average
hazard ratio. *Biometrika*, 68(1), 105-112.

Dormuth, I., Pauly, M., Rauch, G., & Herrmann, C. (2024). Sample size
calculation under nonproportional hazards using average hazard ratios.
*Biometrical Journal*, 66(6), e202300271.

Karrison, T. G. (2016). Versatile tests for comparing survival curves based
on weighted log-rank statistics. *The Stata Journal*, 16(3), 678-690.

Magirr, D., & Ohrn, F. (2026). Robust modestly weighted log-rank tests.
*Pharmaceutical Statistics*, 25(1), e70066.
