---
title: "Validation of FastSurvival"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation of FastSurvival}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
have_survival <- requireNamespace("survival", quietly = TRUE)
have_survRM2  <- requireNamespace("survRM2", quietly = TRUE)
have_survAH   <- requireNamespace("survAH", quietly = TRUE)
have_simtrial <- requireNamespace("simtrial", quietly = TRUE)
```

# Purpose

FastSurvival is built for speed, but a fast estimator is only useful if it
returns the same answer as the established implementation. This vignette
checks the numerical agreement between each FastSurvival function and a
reference from a well-known package: the survival package for the
Kaplan-Meier estimate, the log-rank test, the Cox hazard ratio, and milestone
survival; survRM2 for the restricted mean survival time; survAH for the
average hazard with survival weight; and simtrial for the max-combo test. The
same comparisons form the basis of the automated test suite shipped with the
package.

```{r load}
library(FastSurvival)
```

# Kaplan-Meier survival

`survfit_fast()` evaluates the Kaplan-Meier estimate at a single time point.
We compare against `summary(survfit(...))` at the same time on the `ovarian`
dataset.

```{r km, eval = have_survival}
library(survival)

ord <- order(ovarian$futime)
t_s <- ovarian$futime[ord]
e_s <- ovarian$fustat[ord]

fast <- survfit_fast(t_s, e_s, t_eval = 500, conf.type = "log")

fit  <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
ref  <- summary(fit, times = 500)

data.frame(
  quantity = c("survival", "std.err"),
  fast     = c(unclass(fast)["surv"], unclass(fast)["std.err"]),
  survival = c(ref$surv, ref$std.err),
  row.names = NULL
)
```

# Log-rank test

`survdiff_fast()` returns a two-sided chi-square statistic when `side = 2`,
matching the statistic from `survdiff()`.

```{r logrank, eval = have_survival}
fast_lr <- survdiff_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                         control = 1, side = 2)

ref_lr  <- survdiff(Surv(futime, fustat) ~ rx, data = ovarian)

c(fast = as.numeric(fast_lr), survival = ref_lr$chisq)
```

# Cox hazard ratio

`coxph_fast()` returns the Pike-Halley Estimator, a closed-form approximation
to the Cox partial likelihood maximizer. On data with a moderate hazard ratio
the two agree closely. We report the log hazard ratio from both.

```{r cox, eval = have_survival}
fast_hr <- coxph_fast(ovarian$futime, ovarian$fustat, ovarian$rx, control = 1)

ref_cox <- coxph(Surv(futime, fustat) ~ I(rx == 2), data = ovarian)

c(fast = unclass(fast_hr)["coef"], cox = unname(coef(ref_cox)))
```

# Restricted mean survival time

`rmst_fast()` integrates the Kaplan-Meier survival curve up to a horizon. We
compare the two-group RMST difference against `survRM2::rmst2()`.

```{r rmst, eval = have_survival && have_survRM2}
library(survRM2)

arm  <- as.integer(ovarian$rx == 2)
tau  <- 500

fast_rmst <- rmst_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                       control = 1, tau = tau)

ref_rmst  <- rmst2(time = ovarian$futime, status = ovarian$fustat,
                   arm = arm, tau = tau)

c(fast = unclass(fast_rmst)["diff"],
  survRM2 = ref_rmst$unadjusted.result[1, 1])
```

# Milestone survival

`milestone_fast()` compares Kaplan-Meier survival between two groups at a
milestone timepoint. The per-group survival probabilities match those from
`survfit()`.

```{r milestone, eval = have_survival}
tstar <- 500

fast_ms <- milestone_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                          tau = tstar, method = "wald")

fit_g <- survfit(Surv(futime, fustat) ~ rx, data = ovarian)
ref_g <- summary(fit_g, times = tstar)

ms_surv <- fast_ms$surv

data.frame(
  group    = c("control", "treatment"),
  fast     = c(ms_surv["control"], ms_surv["treatment"]),
  survival = ref_g$surv,
  row.names = NULL
)
```

# Average hazard with survival weight

`ahsw_fast()` computes the average hazard with survival weight of Uno and
Horiguchi. We compare the per-group average hazard against `survAH::ah2()`.

```{r ahsw, eval = have_survival && have_survAH}
library(survAH)

arm <- as.integer(ovarian$rx == 2)
tau <- 500

fast_ah <- ahsw_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                     control = 1, tau = tau)

ref_ah  <- ah2(time = ovarian$futime, status = ovarian$fustat,
               arm = arm, tau = tau)

unclass(fast_ah)[c("ah.ctrl", "ah.trt")]
```

# Max-combo test

`maxcombo_fast()` computes the max-combo test over a set of
Fleming-Harrington weights. We compare the p-value against
`simtrial::maxcombo()` on the same data.

```{r maxcombo, eval = have_survival && have_simtrial}
fast_mc <- maxcombo_fast(ovarian$futime, ovarian$fustat, ovarian$rx,
                         control = 1,
                         rho = c(0, 0, 1, 1), gamma = c(0, 1, 0, 1))

unclass(fast_mc)["p.value"]
```

# Summary

Across all functions the FastSurvival results reproduce the reference values
from the established packages. The point estimates and test statistics agree
to numerical precision for the Kaplan-Meier, log-rank, RMST, milestone, and
average-hazard quantities, and the closed-form Cox hazard ratio agrees with
the partial-likelihood maximizer to the order expected for the Pike-Halley
approximation. This agreement is verified continuously by the package test
suite.

# References

Tang, Y. (2021). Some new confidence intervals for Kaplan-Meier based
estimators from one and two sample survival data. *Statistics in Medicine*,
40(23), 4961-4976.

Uno, H., & Horiguchi, M. (2023). Ratio and difference of average hazard with
survival weight: new measures to quantify survival benefit of new therapy.
*Statistics in Medicine*, 42(7), 936-952.

Karrison, T. G. (2016). Versatile tests for comparing survival curves based
on weighted log-rank statistics. *The Stata Journal*, 16(3), 678-690.
